# Prosper Loan Data Exploration by Peter Carsten Petersen

P4 - Explore and Summarize Data
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(gridExtra)
library(corrplot)
library(PerformanceAnalytics)
library(dplyr)
library(psych)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
loans <- read.csv('prosperLoanData.csv')

loansnew <- loans[, c('Term', 'LoanStatus', 'BorrowerRate', 'EstimatedLoss',  'ProsperRating..Alpha.', 'ProsperScore', 'ListingCategory..numeric.', 'EmploymentStatus', 'IsBorrowerHomeowner', 'CreditScoreRangeLower', 'CreditScoreRangeUpper', 'IncomeRange', 'LoanOriginalAmount')]

loansnew$IncomeRange <- ordered(loansnew$IncomeRange, levels = c("Not displayed", "$0",
                                                          "Not employed" ,"$1-24,999",
                                                          "$25,000-49,999","$50,000-74,999",
                                                          "$75,000-99,999","$100,000+"))

loansnew$ProsperRating..Alpha. <- ordered(loansnew$ProsperRating..Alpha., levels = c('HR','E',
                                                                                     'D','C',
                                                                                     'B','A',
                                                                                     "AA"))

loansnew$CreditScoreRangeAvg <- (loansnew$CreditScoreRangeLower + loansnew$CreditScoreRangeUpper) / 2

loansnew$CreditScoreRangeLower <- NULL
loansnew$CreditScoreRangeUpper <- NULL

loansnew <- subset(loansnew, !is.na(ProsperRating..Alpha.))


```

# Univariate Plots Section
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Overview}
# Overview of data
str(loansnew)

summary(loansnew)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Borrower_rate}
# Histogram Borrower_rate
ggplot(aes(x = BorrowerRate), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420')
  
```



```{r echo=FALSE, message=FALSE, warning=FALSE, Borrower_rate_II}
# Histogram Borrower_rate with adjusted binwidth an x-axis scale
ggplot(aes(x = BorrowerRate), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420', binwidth = 0.02) +
  scale_x_continuous(limits = c(0, 0.35),breaks = seq(0, 0.35, 0.05))
summary(loansnew$BorrowerRate)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, EstimatedLoss}
# Histogram EstimatedLoss
ggplot(aes(x = EstimatedLoss), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420') 


ggplot(aes(x = EstimatedLoss), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420', binwidth = 0.02) +
  xlim(0,0.2)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Term}
# Histogram Term 
ggplot(aes(x = Term), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420')
table(loansnew$Term)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, LoanStatus}
# Bar chart LoanStatus 
ggplot(aes(x = LoanStatus), data = loansnew) +
  geom_bar(color = 'black', fill = '#F79420', stat = 'count') +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```  


```{r echo=FALSE, message=FALSE, warning=FALSE, LoanStatusII}

loansnew$LoanStatus.bucket <- cut(as.numeric(loansnew$LoanStatus), c(1,2,3,4,5,6,12))

levels(loansnew$LoanStatus.bucket) <- c('Chargedoff','Completed','Current','Defaulted','FinalPaymentInProgress','PastDue')


ggplot(aes(x = LoanStatus.bucket), data = loansnew) +
  geom_bar(color = 'black', fill = '#F79420', stat = 'count') +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))


table(loansnew$LoanStatus.bucket)
```





```{r echo=FALSE, message=FALSE, warning=FALSE, ListingCategory}
# Histogram ListingCategory..numeric. 
ggplot(aes(x = ListingCategory..numeric.), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420')
```


```{r echo=FALSE, message=FALSE, warning=FALSE, ListingCategoryincltext}

loansnew$ListingCategory..numeric. <- ordered(loansnew$ListingCategory..numeric.)

levels(loansnew$ListingCategory..numeric.) <- c('Not Available','Debt Consolidation','Home Improvement','Business','Personal Loan','Student Use', 'Auto', 'Other', 'Baby&Adoption','Boat','Cosmetic Procedure','Engagement Ring','Green Loans','Household Expenses','Large Purchases','Medical/Dental','Motorcycle','RV','Taxes','Vacation','Wedding Loans')


ggplot(loansnew,
       aes(x=reorder(ListingCategory..numeric.,ListingCategory..numeric.,
                     function(x)-length(x)))) +
       geom_bar(color = 'black', fill = '#F79420') +
  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(limits = c(0, 55000),breaks = seq(0, 55000, 5000))

prop.table(table(loansnew$ListingCategory..numeric.))
summary(loansnew$ListingCategory..numeric.)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, EmploymentStatus}
# Bar chart EmploymentStatus 
ggplot(aes(x = EmploymentStatus), data = loansnew) +
  geom_bar(color = 'black', fill = '#F79420', stat = 'count') +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```


```{r echo=FALSE, message=FALSE, warning=FALSE, IsBorrowerHomeowner}
# Pie chart IsBorrowerHomeowner 
ggplot(loansnew, aes(x=factor(1), fill= IsBorrowerHomeowner))+
  geom_bar(width = 1, stat = 'count')+
  coord_polar("y")

prop.table(table(loansnew$IsBorrowerHomeowner))

```




```{r echo=FALSE, message=FALSE, warning=FALSE, LoanOriginalAmount}
# Histogram LoanOrginalAmount 
ggplot(aes(x = LoanOriginalAmount), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420', binwidth = 5000)
summary(loansnew$LoanOriginalAmount)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, CreditScoreRangeAvg}
# Histogram CreditScoreRangeAvg 
ggplot(aes(x = CreditScoreRangeAvg), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420')

ggplot(aes(x = CreditScoreRangeAvg), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420', binwidth = 40) +
  scale_x_continuous(breaks = seq(0, 900, 50))

summary(loansnew$CreditScoreRangeAvg)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, IncomeRange}

# Bar chart IncomeRange 
ggplot(aes(x = IncomeRange), data = loansnew) +
  geom_bar(color = 'black', fill = '#F79420', stat = 'count') +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))

```

```{r echo=FALSE, message=FALSE, warning=FALSE, ProsperRating_Score}

# Bar chart ProsperRating..Alpha. 
ggplot(aes(x = ProsperRating..Alpha.), data = loansnew) +
  geom_bar(color = 'black', fill = '#F79420', stat = 'count')
  
# Histogram ProsperScore. 
ggplot(aes(x = ProsperScore), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420', binwidth = 1.5) +
  scale_x_continuous(limits = c(0, 12),breaks = seq(0, 12, 1))

summary(loansnew$ProsperScore)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Borrower_rate_facet}
# Histogram Borrower_rate extended with facet_wrap for incomerange and prosperrating

ggplot(aes(x = BorrowerRate), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420')


ggplot(aes(x = BorrowerRate), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420') +
  facet_wrap(~IncomeRange)



ggplot(aes(x = BorrowerRate), data = loansnew) +
  geom_histogram(color = 'black', fill = '#F79420') +
  facet_wrap(~ProsperRating..Alpha.)

```




# Univariate Analysis

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?



# Bivariate Plots Section
```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
loansnewcorr <- loansnew[, c(1,3,4,5,6,11,12)] 
str(loansnewcorr)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, correlationplot}
loansnewcorr$ProsperRating..Alpha. <- as.numeric(loansnewcorr$ProsperRating..Alpha.)

corrplot(cor(loansnewcorr), method = 'number',tl.col="black", tl.srt=45)

```


```{r echo=FALSE, message=FALSE, warning=FALSE, fullpairsplot}

chart.Correlation(loansnewcorr, histogram=TRUE, pch=19)

```


```{r echo=FALSE, message=FALSE, warning=FALSE, BorroweRatevsProsperRating}

ggplot(aes(x = BorrowerRate, y = ProsperRating..Alpha.), data = loansnew) +
  geom_point()


ggplot(aes(x = ProsperRating..Alpha., y = BorrowerRate), data = loansnew) +
  geom_boxplot()



#describeBy(loansnew$BorrowerRate, loansnew$ProsperRating..Alpha., mat = TRUE)

tapply(loansnew$BorrowerRate, loansnew$ProsperRating..Alpha., summary)


```


```{r echo=FALSE, message=FALSE, warning=FALSE, BorroweRatevsProsperScore}

ggplot(aes(x = BorrowerRate, y = ProsperScore), data = loansnew) +
  geom_point()


ggplot(aes(x = ProsperScore, y = BorrowerRate, group = ProsperScore), data = loansnew) +
  geom_boxplot()

tapply(loansnew$BorrowerRate, loansnew$ProsperScore, summary)

```


```{r echo=FALSE, message=FALSE, warning=FALSE, ProsperRatingvsProsperScore}

ggplot(aes(x = ProsperRating..Alpha., y = ProsperScore), data = loansnew) +
  geom_boxplot()

tapply(loansnew$ProsperScore, loansnew$ProsperRating..Alpha., summary)

```


```{r echo=FALSE, message=FALSE, warning=FALSE, BorroweRatevsLoanOriginalAmount}

ggplot(aes(x = BorrowerRate, y = LoanOriginalAmount), data = loansnew) +
  geom_point()


ggplot(aes(x = BorrowerRate, y = LoanOriginalAmount), data = loansnew) +
  geom_point(alpha = 1/2, size = 3/4, position = 'jitter') +
  scale_x_log10()
  

```

```{r echo=FALSE, message=FALSE, warning=FALSE, BorroweRatevsEstimatedLoss}

ggplot(aes(x = BorrowerRate, y = EstimatedLoss), data = loansnew) +
  geom_point()


ggplot(aes(x = BorrowerRate, y = EstimatedLoss), data = loansnew) +
  geom_point(position = 'jitter') +
  scale_y_continuous(limits = c(0, 0.25),breaks = seq(0, 0.25, 0.05)) +
  scale_x_continuous(limits = c(0, 0.3),breaks = seq(0, 0.3, 0.05)) +
  geom_smooth()


```


```{r echo=FALSE, message=FALSE, warning=FALSE, BorroweRatevsCreditScore}

ggplot(aes(x = BorrowerRate, y = CreditScoreRangeAvg), data = loansnew) +
  geom_point()


ggplot(aes(x = BorrowerRate, y = CreditScoreRangeAvg), data = loansnew) +
  geom_point(alpha = 1/10, size = 1/2, position = 'jitter') +
  geom_smooth()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, EstimatedLossvsProsperRating}

ggplot(aes(x = ProsperRating..Alpha., y = EstimatedLoss), data = loansnew) +
  geom_point()

cor(loansnewcorr$ProsperRating..Alpha., loansnewcorr$EstimatedLoss)
```


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

### What was the strongest relationship you found?




# Multivariate Plots Section

```{r echo=FALSE, ProsperRatingvsProsperScorefaceted}

p1 <- ggplot(aes(y = ProsperScore, x = ProsperRating..Alpha., color = Term), data = loansnew) +
  geom_boxplot()


p2 <- ggplot(aes(y = ProsperScore, x = ProsperRating..Alpha., color = EstimatedLoss), data = loansnew) +
  geom_boxplot()


p3 <- ggplot(aes(y = ProsperScore, x = ProsperRating..Alpha., color = IncomeRange), data = loansnew) +
  geom_boxplot()

grid.arrange(p1, p2, p3, ncol = 1)

ggplot(aes(x = BorrowerRate, y = EstimatedLoss, color = IncomeRange), data = loansnew) +
  geom_line()


```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection
